Generalized Plackett-Luce Likelihoods

### 2022-08-03 inst/jss_hyper3.Rnw
switch to Rnw format for possible inclusion as a vignette
more work

\documentclass[article]{jss}

@@

The \code{hyper2} package~\cite{hankin2017} furnishes computational
support for generalized Plackett-Luce~\citep{plackett1975} likelihood
functions.  Our preferred interpretation is a race (as in track and
field atheletics): given six competitors $1-6$, we ascribe to them
nonnegative strengths $p_1\ldots p_6$; the probability that $i$ beats
$j$ is $p_i/(p_i+p_j)$.  It is conventional to normalise so that the
total strength is unity, and to identify a competitor with his
strength.  Given an order statistic, say $p_1\succ p_2\succ p_3\succ
p_4$, the Plackett-Luce likelihood function would be

@@

The \code{hyper2} package generalized this likelihood function to
functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with

@@

\noindent
where~$\mathcal{O}$ is a set of observations and~$s$ a subset
of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are
integers which may be positive or negative.  The approach adopted by
the~\pkg{hyperdirichlet} package~ is to sptore each of the~$2^n$
possible subsets of~$\left[n\right]$ together with an exponent:

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}

@@

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
preferred interpretation is a race (as in track and field atheletics):
given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1\ldots p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

@@

The \code{hyper2} package generalizes this likelihood function to
functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of observations such as incomplete order
statistics or teams; and the introduction of reified entities allows
one to consider draws~\citep{hankin2009}, noncompetitive tactics such
as collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its parts.

However, consider a race in which there are six runners 1-6 but we
happen to know that three of the runners (1,2,3) are clones of
strength $p_a$, say, two of the runners (4,5) have strength $p_b$, and
the final runner (6) is of strength $p_c$.  The runners race and the
finishing order is:

$$a\succ c\succ b\succ a\succ a \succ b$$

Thus the winner was $a$, second place for $b$, third for $a$, and so
on.  Alternatively we might say that $a$ came first, fourth, and
fifth; $b$ came third and sixth, and $c$ came second.  The
Plackett-Luce likelihood function for $p_a,p_b,p_c$ would be

\begin{equation}
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
\end{equation}

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent this likelihood function, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} used integer-valued sets
together with a print method that used a complicated mapping system
from integers to competitor names.  Current methodology [following
commit \code{51a8b46}] is to use sets of character strings which
represent the competitors directly; this allows for easier combination
of observations including different competitors.}.

<<>>=
2+2
@

\bibliography{chess}
